---
title: "Sample Exploration"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(here)
library(patchwork)
library(ggridges)
```

```{r paths}
if (exists("snakemake")) {
  controls_file <- snakemake@input[["controls"]]
  fair_file     <- snakemake@input[["fair"]]
  clean_us      <- snakemake@input[["us"]]
  clean_ch      <- snakemake@input[["ch"]]
  clean_cn       <- snakemake@input[["cn"]]
  main_text_size <- snakemake@config[["main_text_size"]]
  point_size     <- snakemake@config[["point_size"]]
  line_linewidth <- snakemake@config[["line_linewidth"]]
  colour_scheme  <- snakemake@config[["colour_scheme"]]
} else {
  controls_file <- here("data", "wtc_wtp_controls_tidy.csv")
  fair_file     <- here("data", "wtc_wtp_fair_tidy.csv")
  clean_us      <- here("data", "data_clean_us.csv")
  clean_ch      <- here("data", "data_clean_ch.csv")
  clean_cn      <- here("data", "data_clean_cn.csv")
  main_text_size <- 14
  point_size     <- 3
  line_linewidth <- 1
  colour_scheme  <- "plasma"
}
```

```{r load-data}
edu_num_map <- c(
  no_high_school = 1L, no_matura = 1L,
  high_school    = 2L, matura    = 2L,
  bachelor       = 3L, postgrad  = 4L
)

edu_label_map <- c(
  no_high_school = "Below high school", no_matura = "Below high school",
  high_school    = "High school",       matura    = "High school",
  bachelor       = "Bachelor",          postgrad  = "Postgraduate"
)

edu_levels <- c("Below high school", "High school", "Bachelor", "Postgraduate")

age_levels <- c("18y_24y", "25y_34y", "35y_44y", "45y_54y", "55y_64y", "65y_above")
age_labels <- c(
  "18y_24y" = "18–24", "25y_34y" = "25–34", "35y_44y" = "35–44",
  "45y_54y" = "45–54", "55y_64y" = "55–64", "65y_above" = "65+"
)

df_clean <- bind_rows(
  read_csv(clean_us, show_col_types = FALSE) |> mutate(country = "us"),
  read_csv(clean_ch, show_col_types = FALSE) |> mutate(country = "ch"),
  read_csv(clean_cn, show_col_types = FALSE) |> mutate(country = "cn")
) |>
  select(id, age, education) |>
  mutate(
    education_num   = edu_num_map[education],
    education_label = factor(edu_label_map[education], levels = edu_levels),
    age_num         = as.integer(factor(age, levels = age_levels)),
    age             = factor(age, levels = age_levels)
  )

data_controls <- read_csv(controls_file, show_col_types = FALSE) |>
  filter(!is.na(wtc), time == "pre") |>
  mutate(
    income_decile        = as.integer(income_decile),
    flying_recent_number = replace_na(flying_recent_number, 0),
    country_label        = recode(country, "ch" = "Switzerland", "cn" = "China", "us" = "United States")
  )

data_fair <- read_csv(fair_file, show_col_types = FALSE) |>
  filter(time == "pre") |>
  mutate(
    treatment        = factor(treatment) |> fct_relevel("control"),
    treatment_label  = recode(
      treatment,
      "control" = "Control",       "egal"  = "Egalitarianism",
      "limit"   = "Limitarianism", "prior" = "Prioritarianism",
      "prop"    = "Proportionalism"
    )
  )

df <- data_controls |>
  left_join(df_clean, by = "id") |>
  left_join(
    data_fair |> select(id, fair_self_wtc, fair_group_wtc, fair_self_wtp, fair_group_wtp),
    by = "id"
  )

p95_flights <- quantile(df$flying_recent_number, 0.95, na.rm = TRUE)
```

## Distributions

### Income decile

```{r dist-income}
ggplot(df |> filter(!is.na(income_decile)), aes(x = income_decile)) +
  geom_bar(fill = "#4393C3", alpha = 0.85) +
  facet_wrap(~country_label) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Income decile", y = "Count", title = "Income decile distribution") +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Number of flights

```{r dist-flights}
df |>
  mutate(flights_display = pmin(flying_recent_number, p95_flights)) |>
  ggplot(aes(x = flights_display)) +
  geom_histogram(binwidth = 1, fill = "#74ADD1", alpha = 0.85) +
  facet_wrap(~country_label, scales = "free_y") +
  labs(
    x     = "Number of flights (last 12 months)",
    y     = "Count",
    title = "Number of flights distribution",
    caption = paste0("Capped at 95th percentile (", p95_flights, " flights) for display")
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Climate concern sum score

The sum score combines three items: climate concern for WTC, climate concern for WTP, and an EU-style climate concern item (range 0–15).

```{r dist-clim}
ggplot(df |> filter(!is.na(clim_concern_score)), aes(x = clim_concern_score)) +
  geom_histogram(binwidth = 1, fill = "#66C2A5", alpha = 0.85) +
  facet_wrap(~country_label) +
  scale_x_continuous(breaks = seq(0, 15, by = 3)) +
  labs(
    x     = "Climate concern sum score (0–15)",
    y     = "Count",
    title = "Climate concern sum score distribution"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Education level

Education levels are harmonised across countries: CH "matura"/"no matura" maps to "High school"/"Below high school".

```{r dist-education}
df |>
  filter(!is.na(education_label)) |>
  count(country_label, education_label) |>
  group_by(country_label) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = education_label, y = pct)) +
  geom_col(fill = "#FC8D59", alpha = 0.85) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)), vjust = -0.3, size = 3.5) +
  facet_wrap(~country_label) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(x = "Education level", y = "Share", title = "Education distribution") +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

### Age group

```{r dist-age}
df |>
  filter(!is.na(age)) |>
  count(country_label, age) |>
  group_by(country_label) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = fct_recode(age, !!!age_labels), y = pct)) +
  geom_col(fill = "#FDAE61", alpha = 0.85) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)), vjust = -0.3, size = 3.5) +
  facet_wrap(~country_label) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(x = "Age group", y = "Share", title = "Age distribution") +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Fairness score distributions

Fairness scores reflect respondents' perceptions of the specific fairness framing they were assigned to. Distributions are therefore shown by treatment arm rather than pooled.

```{r dist-fairness, fig.height=6}
data_fair |>
  left_join(
    data_controls |> select(id, country_label),
    by = "id"
  ) |>
  pivot_longer(
    cols      = c(fair_self_wtc, fair_group_wtc, fair_self_wtp, fair_group_wtp),
    names_to  = "fairness_var",
    values_to = "score"
  ) |>
  mutate(
    fairness_label = recode(
      fairness_var,
      fair_self_wtc  = "Personal (WTC)",
      fair_group_wtc = "Group (WTC)",
      fair_self_wtp  = "Personal (WTP)",
      fair_group_wtp = "Group (WTP)"
    ),
    fairness_label = factor(
      fairness_label,
      levels = c("Personal (WTC)", "Group (WTC)", "Personal (WTP)", "Group (WTP)")
    )
  ) |>
  filter(!is.na(score)) |>
  ggplot(aes(x = treatment_label, y = score, fill = treatment_label)) +
  geom_violin(alpha = 0.65, draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary(
    fun = mean, geom = "point", shape = 18, size = point_size, colour = "black"
  ) +
  facet_wrap(~fairness_label, nrow = 1) +
  scale_fill_viridis_d(option = colour_scheme, guide = "none") +
  scale_y_continuous(breaks = 0:5) +
  labs(
    x     = "Treatment arm",
    y     = "Fairness score (0–5)",
    title = "Fairness score distributions by treatment arm",
    caption = "Diamond marks the mean; horizontal lines inside violins are quartiles."
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## Covariate correlations

Spearman correlations are used throughout as the variables are ordinal or right-skewed. Fairness scores are included but note that they are treatment-specific: pooled correlations mix respondents from different arms, which may attenuate or distort relationships with demographic variables.

```{r corr-setup}
vars_corr <- c(
  "income_decile",
  "flying_recent_number",
  "clim_concern_score",
  "fair_self_wtc",
  "fair_group_wtc",
  "fair_self_wtp",
  "fair_group_wtp",
  "education_num",
  "age_num"
)

var_labels_corr <- c(
  income_decile        = "Income decile",
  flying_recent_number = "Nr of flights",
  clim_concern_score   = "Climate concern",
  fair_self_wtc        = "Fairness self (WTC)",
  fair_group_wtc       = "Fairness group (WTC)",
  fair_self_wtp        = "Fairness self (WTP)",
  fair_group_wtp       = "Fairness group (WTP)",
  education_num        = "Education",
  age_num              = "Age"
)

make_corr_heatmap <- function(data, title, label_size = 3.5) {
  mat <- cor(data[, vars_corr], method = "spearman", use = "pairwise.complete.obs")
  as.data.frame(mat) |>
    rownames_to_column("var1") |>
    pivot_longer(-var1, names_to = "var2", values_to = "r") |>
    mutate(
      label1 = factor(var_labels_corr[var1], levels = var_labels_corr),
      label2 = factor(var_labels_corr[var2], levels = rev(var_labels_corr))
    ) |>
    ggplot(aes(x = label1, y = label2, fill = r)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(
      aes(label = sprintf("%.2f", r), colour = abs(r) > 0.45),
      size = label_size
    ) +
    scale_fill_gradient2(
      low = "#4393C3", mid = "white", high = "#D6604D",
      midpoint = 0, limits = c(-1, 1), name = "Spearman ρ"
    ) +
    scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20"), guide = "none") +
    labs(x = NULL, y = NULL, title = title) +
    theme_classic() +
    theme(
      text = element_text(size = main_text_size),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    )
}
```

### Pooled

```{r corr-pooled, fig.width=9, fig.height=7}
make_corr_heatmap(df, "Spearman correlation matrix (pooled)")
```

### By country

```{r corr-by-country, fig.width=18, fig.height=7}
df |>
  group_by(country_label) |>
  group_modify(~ {
    mat <- cor(.x[, vars_corr], method = "spearman", use = "pairwise.complete.obs")
    as.data.frame(mat) |>
      rownames_to_column("var1") |>
      pivot_longer(-var1, names_to = "var2", values_to = "r")
  }) |>
  ungroup() |>
  mutate(
    label1 = factor(var_labels_corr[var1], levels = var_labels_corr),
    label2 = factor(var_labels_corr[var2], levels = rev(var_labels_corr))
  ) |>
  ggplot(aes(x = label1, y = label2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(
    aes(label = sprintf("%.2f", r), colour = abs(r) > 0.45),
    size = 2.8
  ) +
  scale_fill_gradient2(
    low = "#4393C3", mid = "white", high = "#D6604D",
    midpoint = 0, limits = c(-1, 1), name = "Spearman ρ"
  ) +
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20"), guide = "none") +
  facet_wrap(~country_label) +
  labs(x = NULL, y = NULL, title = "Spearman correlation matrix by country") +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Key bivariate relationships

### Income decile × number of flights

```{r bivar-income-flights, fig.height=7}
df |>
  filter(!is.na(income_decile)) |>
  mutate(flights_display = pmin(flying_recent_number, p95_flights)) |>
  ggplot(aes(x = flights_display, y = factor(income_decile), fill = factor(income_decile))) +
  geom_density_ridges(alpha = 0.6, scale = 0.9) +
  facet_wrap(~country_label, nrow = 1) +
  scale_fill_viridis_d(option = colour_scheme, guide = "none") +
  labs(
    x     = "Number of flights (last 12 months)",
    y     = "Income decile",
    title = "Flight frequency by income decile"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Income decile × climate concern

```{r bivar-income-clim}
df |>
  filter(!is.na(income_decile), !is.na(clim_concern_score)) |>
  ggplot(aes(x = factor(income_decile), y = clim_concern_score)) +
  geom_violin(fill = "#66C2A5", alpha = 0.65, draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = point_size) +
  facet_wrap(~country_label, nrow = 1) +
  labs(
    x       = "Income decile",
    y       = "Climate concern sum score",
    title   = "Climate concern by income decile",
    caption = "Diamond marks the mean; lines inside violins are quartiles."
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Number of flights × climate concern

```{r bivar-flights-clim}
df |>
  filter(!is.na(clim_concern_score)) |>
  mutate(flights_display = pmin(flying_recent_number, p95_flights)) |>
  ggplot(aes(x = flights_display, y = clim_concern_score)) +
  geom_point(alpha = 0.12, size = 1) +
  geom_smooth(
    method = "lm", colour = "#D6604D", fill = "#F4A582", linewidth = line_linewidth
  ) +
  facet_wrap(~country_label, nrow = 1) +
  labs(
    x     = "Number of flights (last 12 months)",
    y     = "Climate concern sum score",
    title = "Climate concern by number of flights"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Education × income decile

```{r bivar-educ-income}
df |>
  filter(!is.na(education_label), !is.na(income_decile)) |>
  ggplot(aes(x = education_label, y = income_decile)) +
  geom_violin(fill = "#FC8D59", alpha = 0.65, draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = point_size) +
  facet_wrap(~country_label, nrow = 1) +
  scale_y_continuous(breaks = 1:10) +
  labs(
    x     = "Education level",
    y     = "Income decile",
    title = "Income by education level"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

### Age × climate concern

```{r bivar-age-clim}
df |>
  filter(!is.na(age), !is.na(clim_concern_score)) |>
  ggplot(aes(x = fct_recode(age, !!!age_labels), y = clim_concern_score)) +
  geom_violin(fill = "#FDAE61", alpha = 0.65, draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = point_size) +
  facet_wrap(~country_label, nrow = 1) +
  labs(
    x     = "Age group",
    y     = "Climate concern sum score",
    title = "Climate concern by age group"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = main_text_size),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
